Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2021-11-06 analysis/paper_draft2.Rmd
added keywords; text revisions; facet labels for simplot; revised captions for fig screenshots of app.

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output: 
  rticles::asa_article:
    citation_package: "default"
    keep_tex: TRUE
    #fig_width: 5
    #fig_height: 5
  #output:
#  pdf_document:
#    latex_engine: xelatex
#    number_sections: true
    #fig_width: 4
    #fig_height: 4
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to use historical data on an attorney's use of peremptory strikes in past cases to estimate, in a current trial, how likely an attorney has struck prospective jurors because of cognizable-class membership. We also use the _power prior_ to adjust the weight of such historical data in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with actual historical strike data from a convenience sample of cases from one federal district court.

@@

In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the coverage rates, i.e., the proportion of times we can reach the correct conclusion about the bias parameter based on those credible intervals. That is, when the true value of $b_{curr}$ is greater than zero, the credible interval lies to the right of zero; when $b_{curr}$ equals to zero, the credible interval includes zero; and when $b_{curr}$ is less than zero, the credible interval lies to the left of zero. Since we use a weakly-informative prior that pulls the estimate of $b_{curr}$ towards zero, we focus not on how accurately the model can recover the true value of $b_{curr}$, but on how accurately our model can detect that bias.

@@

For instance, a 90\% coverage rate means that in 900 of 1000 model fits, a given credible interval contained what we fixed as the value of $b_{curr}$. Put another way, if we set $b_{curr}>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero among the 1000 model fits.

@@

```{r fig1, out.width = "95%", fig.cap = "Coverage Rate of bias parameter based on 80 pct credible interval."}

@@

```{r fig2, out.width = "95%", fig.cap = "Coverage Rates - 90 pct. Credible Interval"}

@@

```{r fig3, out.width = "95%", fig.cap = "Coverage Rates - 95 pct. Credible Interval"}

@@

b1.mean <-data.frame(b_s = mean(b1.t$b_s), 
                     `2.5%` =  mean(b1.t$`2.5%`),
                     `97.5%` = mean(b1.t$`97.5%`),
                     sim = "mean", 
                     b = "1", 
                     sig = "Bias",  # FIX: code as function
                     b.num = 1)

b2.mean <-data.frame(b_s = mean(b2.t$b_s), 
                     `2.5%` =  mean(b2.t$`2.5%`),
                     `97.5%` = mean(b2.t$`97.5%`),
                     sim = "mean", 
                     b = "2",
                     sig = "Bias", # FIX: code as function
                     b.num = 2)

@@

simplot <- ggplot(data = b.com, 
                  aes(x = sim.n, y = b_s, color = sig))+
  facet_grid(~b)+
  geom_point(size = 2)+
  geom_errorbar(aes(x = sim.n, ymin = `2.5%`, 
                    ymax = `97.5%`), width = 0.4)+
  geom_hline(data = b.com, aes(yintercept = b.num), col = "red", lwd = 1.1) +
  geom_hline(yintercept = 0, col = "black", lwd = 1, lty = 1) +
  geom_vline(xintercept = 51.25, col = "black", lwd = .5, lty = 2) +
  xlab("Simulation")+
  scale_color_manual("Bias", values = c("blue", "red"))+
  ylab("Posterior Bias Value")+
  ylim(c(-0.5, 5))+
  scale_x_continuous(breaks = c(1:50, 52.5), labels = c(1:50, "Average"))+
  coord_flip()+
  theme_classic()+
  theme(legend.position = "none",
        #axis.title = element_text(size = 16), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 9))

@@

```{r fig4, out.width = "75%", fig.cap="Simulation results for simultaneous strikes example. Columns correspond to values of bias used to simulate data. Points and blue lines denote estimated posterior mean and 95% credible interval for bias parameter. Red line: true value of the bias parameters used in the 50 simulation s."}

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and Shiny [@R; @shiny].

@@

```{r figapp1, out.width='50%', fig.cap = "Density Plots without Historical Strike Data"}

@@

The two vertical dotted lines depict the 80\% credible interval. Here, that credible interval includes zero, indicating no credible to infer bias, given the current strike tally.

@@

```{r figapp2, out.width='50%', fig.cap = "Density Plots with Historical Strike Data (Race)", fig.align='center'}

@@

```{r figapp3, out.width='50%', fig.cap = "Density Plots with Historical Strike Data (Gender)", fig.align='center'}

@@

Finally, if we assign minimal weight to the historical strike data ($a = 0.1$), the density plot updates accordingly (Figure 8).

```{r figapp4, out.width='50%', fig.cap = "Density Plots with Historical Strike Data (Gender, Minimal Weight)", fig.align='center'}

@@

We can find both the prior and posterior density become flatter, and the credible intervals plainly includes zero.

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output: 
  rticles::asa_article:
    citation_package: "default"
    keep_tex: TRUE
    #fig_width: 5
    #fig_height: 5
  #output:
#  pdf_document:
#    latex_engine: xelatex
#    number_sections: true
    #fig_width: 4
    #fig_height: 4
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
keywords: "batson challenge, peremptory strikes"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to combine data on an attorney's strike in a current trial with historical data on that attorney's use of peremptory strikes in past cases to estimate how likely an attorney has struck prospective jurors in the current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical data in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with historical strike data from a convenience sample of cases from one court.

@@

In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the coverage rates, i.e., the proportion of times we can reach the correct conclusion about the bias parameter based on those credible intervals. That is, when the true value of $b_{curr}$ is greater than zero, the credible interval lies to the right of zero; when $b_{curr}$ equals to zero, the credible interval includes zero; and when $b_{curr}$ is less than zero, the credible interval lies to the left of zero. Since we use a weakly-informative prior that pulls the estimate of $b_{curr}$ towards zero, we focus not on how accurately the model can recover the true value of $b_{curr}$, but on how accurately our model can detect that bias. For instance, a 90\% coverage rate means that in 900 of 1000 model fits, a given credible interval contained what we fixed as the value of $b_{curr}$. Put another way, if we set $b_{curr}>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero among the 1000 model fits.

@@

@@

```{r fig1, out.width = "95%", fig.cap = "Coverage Rates based on 80 pct credible interval.", fig.align='center'}

@@

```{r fig2, out.width = "95%", fig.cap = "Coverage Rates - 90 pct. Credible Interval", fig.align='center'}

@@

```{r fig3, out.width = "95%", fig.cap = "Coverage Rates - 95 pct. Credible Interval", fig.align='center'}

@@

b1.mean <-data.frame(b_s = mean(b1.t$b_s), 
                     `2.5%` =  mean(b1.t$`2.5%`),
                     `97.5%` = mean(b1.t$`97.5%`),
                     sim = "mean", 
                     b = "1", 
                     sig = "Bias",
                     b.num = 1)

b2.mean <-data.frame(b_s = mean(b2.t$b_s), 
                     `2.5%` =  mean(b2.t$`2.5%`),
                     `97.5%` = mean(b2.t$`97.5%`),
                     sim = "mean", 
                     b = "2",
                     sig = "Bias",
                     b.num = 2)

@@

cust_labeller_b <- function(x) paste0("b = ", x)

simplot <- ggplot(data = b.com, 
                  aes(x = sim.n, 
                      y = b_s, 
                      color = sig
                      )
                  ) +
  facet_grid(~b, 
             labeller = labeller(b = cust_labeller_b)
             ) +
  geom_point(size = 2)+
  geom_errorbar(aes(x = sim.n, ymin = `2.5%`, 
                    ymax = `97.5%`), width = 0.4)+
  geom_hline(data = b.com, aes(yintercept = b.num), col = "red", lwd = 1.1) +
  geom_hline(yintercept = 0, col = "black", lwd = 1, lty = 1) +
  geom_vline(xintercept = 51.25, col = "black", lwd = .5, lty = 2) +
  xlab("Simulation")+
  scale_color_manual("Bias", values = c("blue", "red"))+
  ylab("Posterior Bias Value")+
  ylim(c(-0.5, 5))+
  scale_x_continuous(breaks = c(1:50, 52.5), labels = c(1:50, "Average"))+
  coord_flip()+
  theme_classic()+
  theme(legend.position = "none",
        #axis.title = element_text(size = 16), 
        axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 9))

@@

```{r fig4, out.width = "75%", fig.cap="Simulation results for simultaneous strikes example. Columns correspond to values of bias used to simulate data. Points and horizontal lines denote estimated posterior mean and 95 pct. credible interval for bias parameter. Vertical lines denote true values of the bias parameters used in the 50 simulations.", fig.align='center'}

@@

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny].

@@

```{r figapp1, out.width='75%', fig.cap = "Screenshot of R-Shiny application showing density plots for bias of prosecutor and defense attorney based on data in 'Current strike tally' table. Vertical dotted lines depict 80 pct. credible interval.", fig.align='center'}

@@

Here, the 80\% credible interval includes zero, indicating no credible to infer bias, given the current strike tally.

@@

```{r figapp2, out.width='75%', fig.cap = "Density Plots with Historical Strike Data (Race). Screenshot of R-Shiny application show density plots based on historical and current strike data for prosecutor and defense attorney.", fig.align='center'}

@@

```{r figapp3, out.width='75%', fig.cap = "Density Plots with Historical Strike Data (Gender). Screenshot of R-Shiny application show density plots based on historical and current strike data for prosecutor and defense attorney. Equal weight ($a = 1$) applied to historical and current data.", fig.align='center'}

@@

```{r figapp4, out.width='75%', fig.cap = "Density Plots with Historical Strike Data (Gender). Screenshot of R-Shiny application show density plots based on historical and current strike data for prosecutor and defense attorney. Minimal weight ($a = 0.1$) applied to historical data.", fig.align='center'}

@@

Finally, if we assign minimal weight to the historical strike data ($a = 0.1$), the density plot updates accordingly (Figure 8). Both the prior and posterior density have become flatter and the credible intervals plainly includes zero.
